---
title: "Exploring Data"
author: "Matt Maslow"
format: html
---

# Reading and Tidying the data

```{r}
library(tidyverse)
library(rvest)
```

## Playing with March Madness data from 2024 tournament

### Webscraping the data & Cleaning

```{r}
url <- "https://www.barttorvik.com/#"
```

```{r}
marchMadness <- 
  url %>%
  read_html() %>%
  html_nodes("table") %>%
  html_table(trim = TRUE, header = FALSE) %>%
  .[[1]] %>%
  .[-1,] %>%
  .[-27,] %>%
  .[-52,] %>%
  .[-77,] %>%
  .[-102,] %>%
  .[-127,] %>%
  .[-152,] # these rows had variables in them, so removed them.

marchMadness
```
```{r}
#write_csv(marchMadness, file = "MarchMadness24.csv")
```

```{r}
mm24 <- read_csv("MarchMadness24.csv")

mm24
```


# Dataset 1: For the shiny table

```{r}
mm24Qualified <- mm24[str_detect(mm24$Team, pattern = "❌|✅"), ]
```

```{r}
# write_csv(mm24Qualified, file = "mm24_TournamentTeams.csv")

mm24Qualified <- read_csv("mm24_TournamentTeams.csv")
```

1. **Conf**: Conference - This variable indicates the conference to which a team belongs.
2. **Games_Played**: The number of games played by the team in the season.
3. **AdjOE**: Adjusted Offensive Efficiency - A metric that measures a team's offensive efficiency while accounting for the strength of the opponent.
4. **AdjDE**: Adjusted Defensive Efficiency - A metric that measures a team's defensive efficiency while accounting for the strength of the opponent.
5. **Barthag**: A metric that estimates a team's expected winning percentage based on its adjusted offensive and defensive efficiencies.
6. **EFG%**: Effective Field Goal Percentage - A shooting metric that adjusts for the value of a three-point field goal relative to a two-point field goal.
7. **EFGD%**: Opponent Effective Field Goal Percentage - Similar to EFG%, but measures the opponent's shooting efficiency.
8. **TOR**: Turnover Rate - The percentage of possessions that end in a turnover for the team.
9. **TORD**: Opponent Turnover Rate - The percentage of possessions that end in a turnover for the opponent.
10. **ORB**: Offensive Rebound Rate - The percentage of available offensive rebounds grabbed by the team.
11. **DRB**: Defensive Rebound Rate - The percentage of available defensive rebounds grabbed by the team.
12. **FTR**: Free Throw Rate - The ratio of free throw attempts to field goal attempts for the team.
13. **FTRD**: Opponent Free Throw Rate - Similar to FTR, but for the opponent.
14. **2P%**: Two-Point Field Goal Percentage - The percentage of two-point field goals made by the team.
15. **2P%D**: Opponent Two-Point Field Goal Percentage - Similar to 2P%, but for the opponent.
16. **3P%**: Three-Point Field Goal Percentage - The percentage of three-point field goals made by the team.
17. **3P%D**: Opponent Three-Point Field Goal Percentage - Similar to 3P%, but for the opponent.
18. **3PR**: Three-Point Attempt Rate - The percentage of field goal attempts that are three-pointers for the team.
19. **3PRD**: Opponent Three-Point Attempt Rate - Similar to 3PR, but for the opponent.
20. **Adj T.**: Adjusted Tempo - A metric that estimates the number of possessions per 40 minutes while adjusting for the pace of play and opponent.
21. **WAB**: Wins Above Bubble - A metric that quantifies a team's wins relative to what a bubble team would typically achieve.




## cleaning data 

```{r}
mm24Qualified <- 
  mm24Qualified %>%
  select(-Rec) %>%
  rename(Games_Played = G) %>%
  mutate(Team = as.character(Team)) %>%
  separate(Team, sep = ",", into = c("Team", "Tournament_Status")) %>%
  separate(Team, sep = "(?<=\\D)(?=\\d)",
           into = c("Team", "PlayOff_Seed")) %>%
  separate(PlayOff_Seed, sep = "\\D", into = c("PlayOff_Seed", "rand")) %>%
  select(-rand, -Tournament_Status) %>%
  mutate(Conf = as.factor(Conf),
         PlayOff_Seed = as.integer(PlayOff_Seed))
mm24Qualified
```

"(?<=\\D)(?=\\d)", this means that there's a non digit that is followed by a digit, in this case space before the digit (utilized stringR cheatsheet)

# Dataset 2: For shiny plot

```{r}
mmHist <- read_csv("mmHistorical.csv")
# https://data.world/michaelaroy/ncaa-tournament-results
```
Here is a file containing every NCAA tournament game result since 1985 (when the tournament was expanded to the 64 team bracket). The dataset contains the year, round (1-6), seed of the teams (1-16), region (1-4) and the scores. 35 years of data, 1985-2019

```{r}
mmHist <- 
  mmHist %>%
  rename(`Seed 1` = Seed...5,
         `Score 1` = Score...6,
         `Team 1` = Team...7,
         `Team 2` = Team...8,
         `Score 2` = Score...9,
         `Seed 2` = Seed...10) %>%
  select(-`Region Number`) %>%
  mutate(Round = as.factor(Round),
         `Region Name` = as.factor(`Region Name`)) %>%
  mutate(winner_score = ifelse(`Score 1` > `Score 2`, `Score 1`, `Score 2`),
             winning_team = ifelse(`Score 1` > `Score 2`, `Team 1`, `Team 2`),
             losing_team = ifelse(`Score 1` > `Score 2`, `Team 2`, `Team 1`),
             losing_score = ifelse(`Score 1` > `Score 2`, `Score 2`, `Score 1`),
             scoreDiff = (winner_score-losing_score)) %>%
  arrange(Year)
mmHist
```



# Final project

```{r}
library(tidyverse)
library(plotly)
library(shiny)
library(bslib)

ui <- navbarPage(
  title = "March Madness Analysis",
  fluid = TRUE, theme = bs_theme(preset = "yeti"), position = "fixed-top",
  tabPanel("Statistics",
           icon = icon("table"),
           tags$br(),
           tags$br(),
           titlePanel(tags$h1("2024 Tournament Results", style = "font-weight: bold; text-align: center; color: darkblue; font-size: 40px;")), 
           tags$p("Author: Matt Maslow", style = "font-size: 18px;  text-align: center; color: darkblue;"),
           tags$hr(style = "border-top: 3px solid #ccc;"),
           fluidRow(
             column(4, offset = 3, radioButtons("stat_type", 
                                    label = "Pick a Stat:",
                                    choices = c("Defensive and Opponent Stats", "Offensive Stats"),
                                    selected = "Offensive Stats")),
             column(4, selectizeInput("Team_choice", label = "Select multiple teams:",
                                      choices = mm24Qualified$Team,
                                      multiple = TRUE))),
           tags$hr(style = "border-top: 3px solid #ccc;"),
           verbatimTextOutput("Variable_Description"),
           tags$hr(style = "border-top: 3px solid #ccc;"),
           dataTableOutput("Team_Table")
  ),
  tabPanel("Historical Plots",
           icon = icon("chart-line"),
           tags$br(),
           tags$br(),
           titlePanel(tags$h1("Tournament Results from 1985 to 2019", style = "font-weight: bold; text-align: center; color: darkblue; font-size: 40px;")),
           tags$p("Author: Matt Maslow", style = "font-size: 18px;  text-align: center; color: darkblue;"),
           tags$br(),
           tags$hr(style = "border-top: 3px solid #ccc;"),
           fluidRow(
             column(3, selectInput("round_choice",
                                   label = "Choose a playoff round:",
                                   choices = c("1", "2", "3", "4", "5", "6"),
                                   selected = "1") ),
             column(3, selectInput("region_choice",
                                   label = "Choose a region:",
                                   choices = c("West", "East", "Midest", "Southeast", "Final Four", "Championship", "South", "Southwest"),
                                   selected = "West") ),
             column(3, sliderInput("year_range",
                                   label = "Choose a year range:",
                                   min = 1985, max = 2019,
                                   value = c(1995, 2009), 
                                   sep = "") ),
             column(3, selectInput("y_var_choice",
                                   label = "Choose Varaible to plot on y-axis:",
                                   choices = c("winner_score", "losing_score", "scoreDiff"),
                                   selected = "winner_score") )),
           tags$hr(style = "border-top: 3px solid #ccc;"),
           tags$p("Select a playoff round from 1 to 6 (1 = first round, 2 = second round, 3 = Sweet 16, 4 = Elite 8, 5 = Final Four, 6 = Championship). The visualization updates based on the chosen round, showing teams' regions still in contention. Adjusting the years can enhance clarity. Utilize the interactive features in the top-right corner to compare game data on hover, displaying team names and results simultaneously.", style = "font-size: 14px; text-align: center; color: black;"),
           tags$hr(style = "border-top: 3px solid #ccc;"),
           plotlyOutput("history_plot", width = "100%", height = 600),
           tags$hr(style = "border-top: 3px solid #ccc;"),
           dataTableOutput("hist_table")
  )
)

server <- function(input, output, session) {
  
  observeEvent(input$round_choice, {
    region_choices <- mmHist %>%
      filter(Round == input$round_choice) %>%
      distinct(`Region Name`) %>%
      pull(`Region Name`)
    updateSelectInput(inputId = "region_choice",
                      choices = region_choices)
  })
  
  mm24TableFiltered <- reactive({
    if (input$stat_type == "Defensive and Oppsonent Stats"){
      mm24TableFiltered <- 
        mm24Qualified %>% 
        filter(Team %in% input$Team_choice) %>%
        select(Team, PlayOff_Seed, Conf, Games_Played, Barthag, AdjDE, `EFG%`,
               DRB, FTRD, `2P%D`, `3P%D`, `Adj T.`, WAB) %>%
        arrange(PlayOff_Seed)
    } else {
      mm24TableFiltered <- 
        mm24Qualified %>% 
        filter(Team %in% input$Team_choice) %>%
        select(Team, PlayOff_Seed, Conf, Games_Played, Barthag, AdjOE, `EFGD%`,
               ORB, FTR, `2P%`, `3P%`, `Adj T.`, WAB) %>%
        arrange(PlayOff_Seed)
    }
  })
  
  output$Variable_Description <- renderPrint({
    if (input$stat_type == "Defensive and Opponent Stats"){
      cat("Defensive and Opponent Stats: Variable Description\n",
          "-------------------------------------------------------\n",
          "Team:        | Team name\n",
          "PlayOff_Seed:| Playoff Seed\n",
          "Conf:        | Conference\n",
          "Games_Played:| Number of games played\n",
          "Barthag:     | Power Rating (Chance of beating an average Division I team)\n",
          "AdjDE:       | Adjusted Offensive Efficiency (An estimate of points allowed per 100 possessions had against an average D1 offense)\n",
          "EFG%:        | Effective Field Goal percentage scored\n",
          "DRB:         | Defensive Rebound percentage\n",
          "FTRD:        | Free Throw Rate allowed\n",
          "2P%D:        | Two-Point allowed\n",
          "3P%D:        | Three-Point allowed\n",
          "Adj T.:      | Adjusted Tempo (estimate of tempo (possessions per 40 minutes) a team has against a team that plays at avg D1 tempo)\n",
          "WAB:         | Wins Above Bubble (the cut off b/n making March Madness Tournament and or not\n"
      )
    } else {
      cat("Offensive Stats: Variable Description\n",
          "-------------------------------------------------------\n",
          "Team:        | Team name\n",
          "PlayOff_Seed:| Playoff Seed\n",
          "Conf:        | Conference\n",
          "Games_Played:| Number of games played\n",
          "Barthag:     | Power Rating (Chance of beating an average Division I team)\n",
          "AdjOE:       | Adjusted Offensive Efficiency (An estimate of points scored per 100 possessions has against an average D1 defense)\n",
          "EFGD%:       | Effective Field Goal Percentage Allowed\n",
          "ORB:         | Offensive Rebound percentage\n",
          "FTR:         | Free Throw Rate (Free Throws Attempted per Field Goal Attempted)\n",
          "2P%:         | Two-Point Field Goal Percentage (Two-Point Field Goal Attempts per Field Goal Attempts)\n",
          "3P%:         | Three-Point Field Goal Percentage (Three-Point Field Goal Attempts per Field Goal Attempts)\n",
          "Adj T.:      | Adjusted Tempo (estimate of tempo (possessions per 40 minutes) a team has against a team that plays at avg D1 tempo)\n",
          "WAB:         | Wins Above Bubble (the cut off b/n making March Madness Tournament and or not\n"
      )
    }
  })
  
  output$Team_Table <- 
    renderDataTable({mm24TableFiltered()},
                    options = list(searching = FALSE, paging = FALSE, info = FALSE, scrollX = TRUE)
    )
  
  mmHistPlot <- reactive({
    if (input$region_choice == "West"){
      mmHist <- mmHist %>%
        filter(`Region Name` == "West" & Round == input$round_choice & Year >= input$year_range[1] & Year <= input$year_range[2])
    } else {
      mmHist <-
        mmHist %>%
        filter(Round == input$round_choice & Year >= input$year_range[1] & Year <= input$year_range[2] & `Region Name` == input$region_choice)
    }
  })
  
  output$history_plot <- renderPlotly({
    
    if (input$y_var_choice == "winner_score"){
      
      plt <- mmHistPlot() %>%
        ggplot(aes(Year, winner_score, fill = winning_team, label = losing_score, label2 = losing_team))+
        geom_point() +
        theme_minimal() +
        scale_colour_viridis_c() +
        labs(title = if(input$year_range[1] == input$year_range[2]){
            paste("The", input$year_range[1], "NCAA Historic Tournament Results:\n \t\tRegion:", input$region_choice, "& Round:", input$round_choice)
          }else{
            paste("The", input$year_range[1], "to", input$year_range[2], "NCAA Historic Tournament Results:\n \t\tRegion:", input$region_choice, "& Round:", input$round_choice)
            },
          x = "Year",
          y = input$y_var_choice,
          fill = "Winning Team") +
        scale_x_continuous(breaks = seq(input$year_range[1], input$year_range[2], 5)) +
        theme(legend.position = "none") +
        theme(plot.title = element_text(face = "bold", size = 14))
      ggplotly(plt, tooltip = c("x", "y", "fill", "label", "label2"))
      
    } else if (input$y_var_choice == "losing_score"){
      
      plt <- mmHistPlot() %>%
        ggplot(aes(Year, losing_score, fill = losing_team, label = winner_score, label2 = winning_team))+
        geom_point() +
        theme_minimal() +
        scale_colour_viridis_c() +
        labs(title = if(input$year_range[1] == input$year_range[2]){
            paste("The", input$year_range[1], "NCAA Historic Tournament Results:\n \t\tRegion:", input$region_choice, "& Round:", input$round_choice)
          }else{
            paste("The", input$year_range[1], "to", input$year_range[2], "NCAA Historic Tournament Results:\n \t\tRegion:", input$region_choice, "& Round:", input$round_choice)
            },
          x = "Year",
          y = input$y_var_choice,
          fill = "Losing Team") +
        scale_x_continuous(breaks = seq(input$year_range[1], input$year_range[2], 5)) +
        theme(legend.position = "none") +
        theme(plot.title = element_text(face = "bold", size = 14))
      ggplotly(plt, tooltip = c("x", "y", "fill", "label", "label2"))
      
    } else if (input$y_var_choice == "scoreDiff"){
      
      plt <- mmHistPlot() %>%
        ggplot(aes(Year, scoreDiff, fill = winning_team, label = winner_score, label2 = losing_team, label3 = losing_score))+
        geom_point() +
        theme_minimal() +
        scale_colour_viridis_c() +
        labs(title = if(input$year_range[1] == input$year_range[2]){
            paste("The", input$year_range[1], "NCAA Historic Tournament Results:\n \t\tRegion:", input$region_choice, "& Round:", input$round_choice)
          }else{
            paste("The", input$year_range[1], "to", input$year_range[2], "NCAA Historic Tournament Results:\n \t\tRegion:", input$region_choice, "& Round:", input$round_choice)
            },
          x = "Year",
          y = "Diff in Losing versus winning scores",
          fill = "Winning Team") +
        scale_x_continuous(breaks = seq(input$year_range[1], input$year_range[2], 5)) +
        theme(legend.position = "none") +
        theme(plot.title = element_text(face = "bold", size = 14))
      ggplotly(plt, tooltip = c("x", "y", "fill", "label", "label2", "label3"))
      
    }
    
  })

  output$hist_table <- renderDataTable({
    mmHistPlot() %>% 
      select(Year, Round, `Region Name`, `Seed 1`, `Team 1`, `Score 1`, `Seed 2`, `Team 2`, `Score 2`) %>%
      mutate(Winner = ifelse(`Score 1` > `Score 2`, `Team 1`, `Team 2`))
  }, options = list(searching = FALSE, paging = FALSE, info = FALSE, scrollX = TRUE))
  
}

shinyApp(ui, server)
```




















