---
title: "🏀 March_Madness 🏀"
format: revealjs
---


```{r}
#| echo: false
#| output: false
#| warning: false
```



```{r}
library(tidyverse)
library(knitr)
library(gridExtra)
library(plotly)

mm24Qualified <- read_csv("mm24_TournamentTeams.csv")
mmHist <- read_csv("mmHistorical.csv")
```

```{r}
mm24Qualified <- 
  mm24Qualified %>%
  select(-Rec) %>%
  rename(Games_Played = G) %>%
  mutate(Team = as.character(Team)) %>%
  separate(Team, sep = ",", into = c("Team", "Tournament_Status")) %>%
  separate(Team, sep = "(?<=\\D)(?=\\d)",
           into = c("Team", "PlayOff_Seed")) %>%
  separate(PlayOff_Seed, sep = "\\D", into = c("PlayOff_Seed", "rand")) %>%
  select(-rand, -Tournament_Status) %>%
  mutate(Conf = as.factor(Conf),
         PlayOff_Seed = as.integer(PlayOff_Seed))

mmHist <- 
  mmHist %>%
  rename(`Seed 1` = Seed...5,
         `Score 1` = Score...6,
         `Team 1` = Team...7,
         `Team 2` = Team...8,
         `Score 2` = Score...9,
         `Seed 2` = Seed...10) %>%
  select(-`Region Number`) %>%
  mutate(Round = as.factor(Round),
         `Region Name` = as.factor(`Region Name`)) %>%
  mutate(winner_score = ifelse(`Score 1` > `Score 2`, `Score 1`, `Score 2`),
             winning_team = ifelse(`Score 1` > `Score 2`, `Team 1`, `Team 2`),
             losing_team = ifelse(`Score 1` > `Score 2`, `Team 2`, `Team 1`),
             losing_score = ifelse(`Score 1` > `Score 2`, `Score 2`, `Score 1`),
             scoreDiff = (winner_score-losing_score)) %>%
  arrange(Year)
```



## Data

### March Madness 2024 Qualified Teams

 - There are 64 teams that qualified for the 2024 March Madness tournament. 
 - Data consist of 24 variables, half offensive stats and half defensive stats. 
 - Will be using this data to analyze why some teams are stronger and were more likely of having greater success in the tournament.

### Historical Data

 - This data contains historical data from years 1985 to 2019, with all six rounds of the playoffs including team names and scores. 
 -Will be using this data to analyze the winning scores and score differences in the final round of the playoffs, and to see if there were any record breaking scores.


## Vizualization 1 (Table)

```{r}
mm24Qualified %>% 
  filter(Games_Played >= 37) %>%
        select(Team, PlayOff_Seed, Conf, Games_Played, Barthag, AdjOE, `EFGD%`,
               ORB, FTR, `2P%`, `3P%`, `Adj T.`, WAB) %>%
        arrange(PlayOff_Seed) %>%
        kable(sortable = TRUE, bordered = TRUE, highlight = TRUE, width = "100%")
```


## Vizualization 2 (Plot)

When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:
```{r}
plt1 <- mmHist %>%
  filter(Year >= 1990 & Year <= 2000 & Round == "6") %>%
  ggplot(aes(Year, winner_score, fill = winning_team, label = losing_score, label2 = losing_team)) +
  geom_point() +
  theme_minimal() +
  scale_colour_viridis_c() +
  labs(x = "Year",
  y = "Winning Score",
  fill = "Winning Team") +
  scale_x_continuous(breaks = seq(1990, 2000, 5)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(face = "bold", size = 14))
plt1 <- ggplotly(plt1, tooltip = c("x", "y", "fill", "label", "label2"))

plt3 <- mmHist %>%
  filter(Year >= 1990 & Year <= 2000 & Round == "6") %>%
  ggplot(aes(Year, scoreDiff, fill = winning_team, label = winner_score, label2 = losing_team, label3 = losing_score))+
  geom_point() +
  theme_minimal() +
  scale_colour_viridis_c() +
  labs(title = "Winning Score(Top) & Score Difference(Bottom)
       in Final Round",
  x = "Year",
  y = "Score Difference",
  fill = "Winning Team") +
  scale_x_continuous(breaks = seq(1990, 2000, 5)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(face = "bold", size = 14))
plt3 <- ggplotly(plt3, tooltip = c("x", "y", "fill", "label", "label2", "label3"))
subplot(plt1, plt3, nrows = 2)
```


